Incremental tactics

### 2012-05-08 doc/papers/LFMTP2012/paper.tex
paper

\subsection{Core calculus}

The core calculus defines the \emph{values} we want to
manipulate. These contain constants, and --- contrarily to \eg\
\sysname{ML} values --- binders and variables enabling $\lambda$-tree
syntax. Thus, they embed a superficial notion of computation:
hereditary substitution. This calculus is an extension of \emph{spine
  canonical \LF} \cite{pfenning2007term} enriched with metavariables
which are references to open terms as in
\cite{nanevski2008contextual}. We present its typing algorithm
relatively to a signature. It features an explicit caching mechanism
using metavariables.

@@

\begin{align*}
  K &\gequal
  \prd x A K \gor
  \type &
  \text{Kind}\\
  A &\gequal
  \prd x A A \gor
  \app {\cst a} S &
  \text{Type family} \\
  M &\gequal
  \lam x M \gor
  F \gor
  \smeta X \sigma &
  \text{Object} \\
  F &\gequal \app H S &
  \text{Application} \\
  H &\gequal
  \var x \gor
  \cst c &
  \text{Head}\\
  S &\gequal
  \spinenil \gor
  \spinecons M S &
  \text{Spine}\\
  \sigma &\gequal
  \msubstnil \gor
  \msubstcons \sigma x M &
  \text{Parallel substitution} \\
  \Gamma &\gequal
  \envnil \gor
  \envcons \Gamma x A &
  \text{Environment}
\end{align*}

@@

Metavariables $\smeta X \sigma$ stand for open objects with free
variables in $\dom{\sigma}$; they are defined in a repository (see
below). We write $\FMV{M}$ for the set of metavariables in $M$ and
speak of a \emph{metaclosed} object $M$ if $\FMV{M} = \emptyset$.

@@

A \emph{repository} $\mr$ is a finite map from metavariable to
judgements of object well-typing, along with a distinguished
metavariable, the \emph{head}:
$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp M A)), \smeta X \sigma $$
We maintain the invariant that a repository $\mr$ is
\emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint
of $\FMV{\meta X}$ does not contain $\meta X$. With $\mr =
(\Delta, X[\sigma])$, we write $\hat {\mr}$ for $X[\sigma]$,
abusively $\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for
$\gsubst M\sigma$ if $\Delta(\meta X) = (\Gamma\vdash M : A)$.

@@

    % App
    \infer{
      \jf\mr\Gamma F \mmr {M'} {A'}
       \and
      \jea\mmr\Gamma {\app{\cst a} S} {A'}
    }{
      \jm\mr\Gamma F {\app{\cst a} S} \mmr {M'}
    }
    % TODO commenter type ground

    % Meta
    \infer{
      \mr(\meta X) = (\Gamma'\vdash M:A')
      \and
      \js\mr\Gamma \sigma {\Gamma'} \mmr {\sigma'}
      \and
      \jea\mmr\Gamma A {\gsubst {A'} {\sigma'}}
    }{
      \jm\mr\Gamma {\smeta X \sigma} A \mmr {M'}
    }
  \end{mathpar}

@@

  \fbox{$\jf\mr\Gamma F \mmr M A$}
  \qquad
  {Application}

@@

    % Sliceable
    \infer{
      \jh\mr\Gamma H A \annsliceable
      \and
      \jl\mr\Gamma A S \mmr {S'} {A'}
      \and
      \jp\mmr\Gamma {\app H S} A \mmmr
    }{
      \jf\mr\Gamma {\app H S} \mmmr {\hat\mmmr} {A'}
    }

@@

  \fbox{$\jp\mr\Gamma M A \mmr$}
  \qquad
  {Push}

  \begin{mathpar}
    \infer{
      \stren{\Gamma, M}=\Gamma'
      \and
      \text{$\meta X$ fresh in $\Delta$}
    }{
      \jp{(\Delta; \smeta Z {\sigma})}\Gamma M A {(\mapadd\Delta{\meta
          X}{(\Gamma'\vdash M:A)}; \smeta X {\msubstid\Gamma})}
    }
  \end{mathpar}

@@

The main judgement is application typing $\jf\mr\Gamma F \mmr M A$,
which means that in repository $\mr$ and environment $\Gamma$, the
applicative term $F$ is well-typed of type $A$, and the slicing of $F$
produces a new repository $\mmr$. The object $M$ is the
\emph{residual} of this operation: only certain kind of applications
are sliced, and it is possible that the ``tip'' of application $F$ is
not sliced (because it is declared as non-sliceable). In that case,
$M$ is this tip; if $F$ is directly sliceable, then $M$ is the
metavariable referring to it.

Definitional equality judgement between two families $\jea\mr\Gamma A
{A'}$ is used by these typing rules. This equivalence is usually
defined to be just $\alpha$-equivalence since objects are in
$\beta\eta$-canonical form. Here, objects can contain metavariables
defined in $\mr$, so we shall thread it. Moreover, we will see in the
next section how defined constants compute and thus influence
definitional equality. Definitional equality is defined later in Fig.
\ref{fig:def-eq}.

>>>>

\subsection{Value calculus}

The value calculus defines the \emph{values} we want to
manipulate. These contain constants, and --- contrarily to \eg\
\sysname{ML} values --- binders and variables enabling $\lambda$-tree
syntax. Thus, they embed a superficial notion of computation:
hereditary substitution. This calculus is an extension of \emph{spine
  canonical \LF} \cite{pfenning2007term} enriched with metavariables
which are references to open terms as in
\cite{nanevski2008contextual}. We present its typing algorithm
relatively to a signature. It features an explicit caching mechanism
using metavariables.

@@

\begin{align*}
  K &\gequal
  \prd x A K \gor
  \type &
  \text{Kind}\\
  A &\gequal
  \prd x A A \gor
  \app {\cst a} S &
  \text{Type family} \\
  M &\gequal
  \lam x M \gor
  F &
  \text{Object} \\
  F &\gequal \app H S
  \gor
  \smeta X \sigma &
  \text{Application} \\
  H &\gequal
  \var x \gor
  \cst c &
  \text{Head}\\
  S &\gequal
  \spinenil \gor
  \spinecons M S &
  \text{Spine}\\
  \sigma &\gequal
  \msubstnil \gor
  \msubstcons \sigma x M &
  \text{Parallel substitution} \\
  \Gamma &\gequal
  \envnil \gor
  \envcons \Gamma x A &
  \text{Environment}
\end{align*}

@@

% TODO com M check/F infer

Metavariables $\smeta X \sigma$ stand for open objects with free
variables in $\dom{\sigma}$; To each use of them is attached a
substitution $\sigma$ defining these free variables. They are defined
in a repository (see below). We write $\FMV{M}$ for the set of
metavariables in $M$ and speak of a \emph{metaclosed} object $M$ if
$\FMV{M} = \emptyset$.

@@

A \emph{repository} $\mr$ is a finite map from metavariable to
judgements of object well-typing, along with a distinguished
metavariable, the \emph{head}:
$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp M A)), \smeta X \sigma $$
We maintain the invariant that a repository $\mr$ is
\emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint
of $\FMV{\meta X}$ does not contain $\meta X$. If $\mr =
(\Delta, X[\sigma])$, we write $\hat {\mr}$ for $X[\sigma]$,
abusively $\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for
$\gsubst M\sigma$ if $\Delta(\meta X) = (\Gamma\vdash M : A)$.

@@

    % App
    \infer{
      \jf\mr\Gamma F \mmr {F'} {A'}
       \and
      \jea\mmr\Gamma {\app{\cst a} S} {A'}
    }{
      \jm\mr\Gamma F {\app{\cst a} S} \mmr {F'}
    }
    % TODO commenter type ground
  \end{mathpar}

@@

  \fbox{$\jf\mr\Gamma F \mmr F A$}
  \qquad
  {Application}

@@

    % Meta
    \infer{
      \mr(\meta X) = (\Gamma'\vdash M:A)
      \and
      \js\mr\Gamma \sigma {\Gamma'} \mmr {\sigma'}
    }{
      \jf\mr\Gamma {\smeta X \sigma} \mmr {\smeta X {\sigma'}} {\gsubst{A}\sigma'}
    }

    % Sliceable
    \infer{
      \jh\mr\Gamma H A \annsliceable
      \and
      \jl\mr\Gamma A S \mmr {S'} {A'}
      \and
      \stren{\Gamma, \app H {S'}}=\Gamma'
      \and
      \text{$\meta X$ fresh in $\mr$}
    }{
      \jf\mr\Gamma {\app H S} {\mapadd\mmr{\meta X}{(\Gamma'\vdash
          {\app H {S'}}:A)}} {\smeta X{\msubstid{\Gamma'}}} {A'}
    }

@@

@@

The main judgement is application typing $\jf\mr\Gamma F \mmr {F'} A$,
which means that in repository $\mr$ and environment $\Gamma$, the
applicative term $F$ is well-typed of type $A$, and the slicing of $F$
produces a new repository $\mmr$. The object $F'$ is the
\emph{residual} of this operation: only certain kind of applications
are sliced, and it is possible that the ``tip'' of application $F$ is
not sliced (because it is declared as non-sliceable). In that case,
$M$ is this tip; if $F$ is directly sliceable, then $M$ is the
metavariable referring to it.

These typing rules use definitional equality judgement between two
families $\jea\mr\Gamma A {A'}$. This equivalence is usually defined
to be just $\alpha$-equivalence since objects are in
$\beta\eta$-canonical form. Here, objects can contain metavariables
defined in $\mr$, so we shall thread it. Moreover, we will see in the
next section how defined constants compute and thus influence
definitional equality. Definitional equality is defined later in Fig.
\ref{fig:def-eq}.
